Towards a Classification of Isolated $j$-invariants

### 2023-07-27 TeX/draft.tex
Small edits

\newcommand{\davidnote}[1]{\textcolor{Plum}{David: #1}}
\newcommand{\filip}[1]{{\textcolor{cyan}{Filip: [#1]}}}
\newcommand{\abbey}[1]{\textcolor{blue}{Abbey: #1}}
\newcommand{\dld}[1]{\textcolor{Plum}{dld: #1}}
\newcommand{\sachi}[1]{\textcolor{purple}{Sachi: #1}}
\newcommand{\timo}[1]{\textcolor{red}{Timo: #1}}
\newcommand{\travis}[1]{\textcolor{orange}{Travis: #1}}
\newcommand{\abedit}[1]{{\color{blue} #1}}
\newcommand{\fedit}[1]{{\color{cyan} #1}}
\newcommand{\sedit}[1]{{\color{purple} #1}}

@@

\begin{document}
\begin{abstract} 
We develop an algorithm to test whether a non-CM elliptic curve $E/\Q$ gives rise to an isolated point of any degree on any modular curve of the form $X_1(N)$. This builds on prior work of Zywina which gives a method for computing the image of the adelic Galois representation associated to $E$. Running this algorithm on all elliptic curves presently in the $L$-functions and Modular Forms Database yields strong evidence for the conjecture that $E$ gives rise to a $\PP^1$-isolated point on $X_1(N)$ if and only if $j(E)=-140625/8, -9317,$ $351/4$, or $-162677523113838677$.
    \end{abstract}
\maketitle

@@

The modular curve $X_1(N)$ is an algebraic curve over $\Q$ whose non-cuspidal points parametrize elliptic curves with a distinguished point of order $N$. We are interested in studying isolated points on $X_1(N)$. Roughly speaking these are points of degree $d$ not belonging to an infinite family of degree $d$ points parametrized by a geometric object. For example, if $f: X_1(N) \rightarrow \PP^1$ is a rational map of degree $d$, then $f^{-1}(\PP^1(\Q))$ contains infinitely many closed points of degree $d$ by Hilbert's irreducibility theorem \cite[Chapter~9]{serre97}. We say a degree $d$ point \emph{not} arising from such a map is \textbf{$\PP^1$-isolated}. Other infinite families of degree $d$ points correspond to positive rank abelian subvarieties of the curve's Jacobian; see Section \ref{sec:background} for details. A point which is not thus parametrized is \textbf{AV-isolated}. If a closed point $x\in X_1(N)$ is both $\PP^1$- and AV-isolated, then we say $x$ is \textbf{isolated}. Isolated points can be thought of as those points whose existence is not explained by a geometric construction. The lack of control of isolated points in infinite families of modular curves lies at the heart of several important open problems. See, for example, \cite[Question~1]{MazurNote} and \cite[Theorem~1.3]{BourdonNajman2021}.

One special class of isolated points is \textbf{sporadic} points, which are $x \in X_1(N)$ such that there are only finitely many points on $X_1(N)$ of degree at most $\deg(x)$. Hence a non-cuspidal sporadic point corresponds to an elliptic curve with a point of order $N$ in ``usually low degree." Elliptic curves with complex multiplication (CM) provide many natural examples of sporadic points, since the extra endomorphisms of a CM elliptic curve constrain the size of the image of the associated Galois representation. Indeed, as shown in \cite[Theorem~8.2]{CGPS2022}, there exist sporadic CM points on $X_1(N)$ for all $N\geq 721$.

@@

\noindent Since the image of the adelic Galois representation of a non-CM elliptic curve $E/\Q$ has finite level, this theorem reduces the problem of determining whether $j(E)$ is isolated to the problem of determining whether $E$ gives rise to an isolated point on a finite number of modular curves. While a sufficient finite list can be deduced directly from the adelic level as found by Zywina \cite{ZywinaAlgorithm}, our output works to reduce this list to its minimal size. This inspires the notion of \textbf{primitive points} on modular curves, as studied in Section \ref{sec:PrimitivePoints}. In particular, $j(E)$ is isolated if and only if there exists an isolated primitive point corresponding to $E$. Moreover, any point $x\in X_1(N)$ with $j(x)=j(E)$ corresponds to a \emph{unique} primitive point (Proposition \ref{Prop:single_sink}).

@@

\section*{Acknowledgments}
We thank Pete Clark, Maarten Derickx, Jeremy Rouse, and David Zureick-Brown for helpful conversations.

>>>>

\newcommand{\davidnote}[1]{\textcolor{Plum}{David: #1}}
\newcommand{\filip}[1]{{\textcolor{cyan}{Filip: [#1]}}}
\newcommand{\abbey}[1]{\textcolor{blue}{Abbey: #1}}
\newcommand{\dld}[1]{\textcolor{Plum}{dld: #1}}
\newcommand{\sachi}[1]{\textcolor{purple}{Sachi: #1}}
\newcommand{\timo}[1]{\textcolor{red}{Timo: #1}}
\newcommand{\travis}[1]{\textcolor{orange}{Travis: #1}}
\newcommand{\dldedit}[1]{{\color{Plum} #1}}
\newcommand{\abedit}[1]{{\color{blue} #1}}
\newcommand{\fedit}[1]{{\color{cyan} #1}}
\newcommand{\sedit}[1]{{\color{purple} #1}}

@@

\begin{document}

\begin{abstract}
We develop an algorithm to test whether a non-CM elliptic curve $E/\Q$ gives rise to an isolated point of any degree on any modular curve of the form $X_1(N)$. This builds on prior work of Zywina which gives a method for computing the image of the adelic Galois representation associated to $E$. Running this algorithm on all elliptic curves presently in the $L$-functions and Modular Forms Database yields strong evidence for the conjecture that $E$ gives rise to a $\PP^1$-isolated point on $X_1(N)$ if and only if $j(E)=-140625/8, -9317,$ $351/4$, or $-162677523113838677$.
\end{abstract}

\maketitle

@@

The modular curve $X_1(N)$ is an algebraic curve over $\Q$ whose non-cuspidal points parametrize elliptic curves with a distinguished point of order $N$. We are interested in studying isolated points on $X_1(N)$.
Roughly speaking, these are points of degree $d$ not belonging to an infinite family of degree $d$ points parametrized by a geometric object. For example, if $f: X_1(N) \rightarrow \PP^1$ is a rational map of degree $d$, then $f^{-1}(\PP^1(\Q))$ contains infinitely many closed points of degree $d$ by Hilbert's irreducibility theorem \cite[Chapter~9]{serre97}. We say a degree $d$ point \emph{not} arising from such a map is \textbf{$\PP^1$-isolated}. Other infinite families of degree $d$ points correspond to positive rank abelian subvarieties of the curve's Jacobian; see Section \ref{sec:background} for details. A point which is not thus parametrized is \textbf{AV-isolated}. If a closed point $x\in X_1(N)$ is both $\PP^1$- and AV-isolated, then we say $x$ is \textbf{isolated}.

Isolated points can be thought of as those points whose existence is not explained by a geometric construction. The lack of control of isolated points in infinite families of modular curves lies at the heart of several important open problems. See, for example, \cite[Question~1]{MazurNote} and \cite[Theorem~1.3]{BourdonNajman2021}.

One special class of isolated points is \textbf{sporadic} points, which are
points $x \in X_1(N)$ such that there are only finitely many points on $X_1(N)$
of degree at most $\deg(x)$. Hence a non-cuspidal sporadic point corresponds to
an elliptic curve with a point of order $N$ \dldedit{with ``unusually low degree.''}
Elliptic curves with complex multiplication (CM) provide many natural examples
of sporadic points, since the extra endomorphisms of a CM elliptic curve
constrain the size of the image of the associated Galois representation.
Indeed, as shown in \cite[Theorem~8.2]{CGPS2022}, there exist sporadic CM
points on $X_1(N)$ for all $N\geq 721$.

@@

\noindent Since the image of the adelic Galois representation of a non-CM elliptic curve $E/\Q$ has finite level, this theorem reduces the problem of determining whether $j(E)$ is isolated to the problem of determining whether $E$ gives rise to an isolated point on a finite number of modular curves.
While a sufficient finite list can be deduced directly from the adelic level as found by Zywina's Algorithm~\cite{ZywinaAlgorithm}, our \dldedit{approach attempts to reduce the size of this list as much as possible}.
This inspires the notion of \textbf{primitive points} on modular curves, as studied in Section \ref{sec:PrimitivePoints}. In particular, $j(E)$ is isolated if and only if there exists an isolated primitive point corresponding to $E$. Moreover, any point $x\in X_1(N)$ with $j(x)=j(E)$ corresponds to a \emph{unique} primitive point (Proposition \ref{Prop:single_sink}).

@@

\section*{Acknowledgments}
We thank Pete Clark, Maarten Derickx, Jeremy Rouse, and David Zureick-Brown for helpful conversations.
\dld{Perhaps we also give an explicit shoutout to Zywina for making his code public too?}
